Refinement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-09-20 rigid_bodies.tex
Replacement of \text{\tiny B} by \mini{B}

In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j = 1, \cdots, M$.
The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an \textit{extended energy} given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

@@

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of Tuckerman \textit{et al}. \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

@@

By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of Sec.~\ref{sec:nve} and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
This one can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence: it ensures
the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation}
\label{eq:equipartition}
L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
\end{equation}
where subscript $\timestep$ points out that this average depends on the employed time-step size.
The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
We emphasize that the temperature in Eq.~\eqref{eq:equipartition} concerns the system that has actually been simulated, rather than the one initially specified.

@@

Bringing the attention back to the NVT case, our proposal for dealing with discretization effects consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
The only required modification is that $\refined K$ should replace $K$ in Eq.~\eqref{eq:generalized force NVT}, which then becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\text{\tiny B} T.
\end{equation}

Also, the \textit{refined extended energy} preserved by this approach is now given by
\begin{equation}
\label{eq:this extended energy}
{\refined H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

@@

Finally, we recall that configurations are sampled from a distribution proportional to $e^{-\beta \refined{\Ham{}}}$.
Notwithstanding, it is possible to compute ensemble averages consistent with a distribution proportional to $e^{-\beta \Ham{}}$, which would have been obtained if $\timestep \to 0$.
For this, we simply need to reweight any computed observable $A$ in accordance with \cite{Torrie_1977}
\begin{equation}
\label{eq:reweighting}
\langle A \rangle_0 = \frac{\left\langle A \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}{\left\langle \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}.
\end{equation}

>>>>

In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j = 1, \cdots, M$.
The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an \textit{extended energy} given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\mini{B} T\eta_1 + k_\mini{B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\mini{B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\mini{B} T t_d^2$ and $Q_j = k_\mini{B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

@@

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\mini{B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\mini{B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\mini{B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of Tuckerman \textit{et al}. \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

@@

By defining a propagator $e^{\timestep \mathcal{L}_\mini{NVT}}$, where $\mathcal{L}_\mini{NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\mini{NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\mini{NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\mini{NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of Sec.~\ref{sec:nve} and $\mathcal{L}_\mini{NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
This one can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence: it ensures
the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation}
\label{eq:equipartition}
L k_\mini{B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
\end{equation}
where subscript $\timestep$ points out that this average depends on the employed time-step size.
The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
We emphasize that the temperature in Eq.~\eqref{eq:equipartition} concerns the system that has actually been simulated, rather than the one initially specified.

@@

Bringing the attention back to the NVT case, our proposal for dealing with discretization effects consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\mini{B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
The only required modification is that $\refined K$ should replace $K$ in Eq.~\eqref{eq:generalized force NVT}, which then becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\mini{B} T.
\end{equation}

Also, the \textit{refined extended energy} preserved by this approach is now given by
\begin{equation}
\label{eq:this extended energy}
{\refined H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\mini{B} T\eta_1 + k_\mini{B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

@@

Finally, we recall that configurations are sampled from a distribution proportional to $e^{-\beta \refined{\Ham{}}}$.
Notwithstanding, it is possible to compute ensemble averages consistent with a distribution proportional to $e^{-\beta \Ham{}}$, which would have been obtained if $\timestep \to 0$.
For this, we simply need to reweight any computed observable $A$ in accordance with \cite{Torrie_1977}
\begin{equation}
\label{eq:reweighting}
\langle A \rangle_0 = \frac{\left\langle A \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\mini{B} T}}\right)\right\rangle_\timestep}{\left\langle \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\mini{B} T}}\right)\right\rangle_\timestep}.
\end{equation}
